An Improved Method for Coupling Hydrodynamics with Reaction Networks and Nuclear Statistical Equilibrium

### 2021-09-26 simplified_sdc_intro/paper.tex
update
some restructuring of the first 2 sections

\newcommand{\nablab}{{\mathbf{\nabla}}}
\newcommand{\Ub}{\mathbf{U}}
\newcommand{\gb}{\mathbf{g}}
\newcommand{\omegadot}{\dot{\omega}}
\newcommand{\Sdot}{\dot{S}}
\newcommand{\ddx}[1]{{\frac{{\partial#1}}{\partial x}}}
\newcommand{\ddt}[1]{{\frac{{\partial#1}}{\partial t}}}
\newcommand{\odt}[1]{{\frac{{d#1}}{dt}}}
\newcommand{\divg}[1]{{\nablab \cdot \left (#1\right)}}
\newcommand{\dedr}{\left . {\partial{}e}/{\partial\rho}\right |_{T, X_k}}
\newcommand{\dedrd}{\left . \frac{\partial{}e}{\partial\rho}\right |_{T, X_k}}
\newcommand{\dedX}{\left . {\partial{}e}/{\partial{}X_k} \right |_{\rho, T}}
\newcommand{\dedXd}{\left . \frac{\partial{}e}{\partial{}X_k} \right |_{\rho, T, X_{j,j\ne k}}}
\newcommand{\dedT}{\left . {\partial{}e}/{\partial{}T} \right |_{\rho,X_k}}
\newcommand{\dedTd}{\left . \frac{\partial{}e}{\partial{}T} \right |_{\rho,X_k}}

@@

%% \author[0000-0001-8401-030X]{M.~Zingale}
%% \affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
%%        Stony Brook, NY 11794-3800}

%% \author[0000-0003-0439-4556]{M.~P.~Katz}
%% \affiliation{Nvidia Corp}

%% \author[0000-0002-5749-334X]{J.~B.~Bell}
%% \affiliation{Center for Computational Sciences and Engineering, Lawrence Berkeley National Laboratory, Berkeley, CA  94720}

%% \author[0000-0003-1791-0265]{A.~J.~Nonaka}
%% \affiliation{Center for Computational Sciences and Engineering, Lawrence Berkeley National Laboratory, Berkeley, CA  94720}

%% \author[0000-0001-8092-1974]{W.~Zhang}
%% \affiliation{Center for Computational Sciences and Engineering, Lawrence Berkeley National Laboratory, Berkeley, CA  94720}

%% \correspondingauthor{Michael Zingale}
%% \email{michael.zingale@stonybrook.edu}

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
Reacting astrophysical flows can be challenging to model because of
the difficulty in accurately coupling hydrodynamics and reactions.
This can be particularly acute at high temperatures where nuclear
statistical equilibrium is established.  We develop a simplified
approach based on spectral deferred corrections (SDC) coupling of explicit
hydrodynamics and stiff reaction sources as an alternative to operator
splitting or the more comprehensive SDC approach we demonstrated
previously.  We apply the new method to some example problems and show
how to modify it to work with a hybrid network consisting of a
reaction ODE system and a table for nuclear statistical equilibrium.
This is all done in the framework of the \castro\ hydrodynamics code,
and all algorithm implementations are freely available.
\end{abstract}

@@

Reactive flow can be challenging to model when changes in the
nuclear abundances due to reactions are integrated over hydrodynamic
time scales.  Traditional methods of coupling
hydrodynamics and reactions used in astrophysics use operator
splitting---each physical process acts on the output of the previous
process in alternating fashion.  This makes it easy to add physics in a modular way to a
simulation code, but competition between physical processes can cause
the coupling to breakdown.  These splitting errors can lead to loss of
accuracy and further time step limitations.

@@

In \cite{castro_sdc}, we developed second- and fourth-order accurate
methods in space and time for coupling hydrodynamics and nuclear
reaction networks based on spectral deferred corrections (SDC)
methods, and demonstrated the method in a variety of test problems.
The approach presented here is considerably simpler from a temporal
integration standpoint, but allows is to incorporate modifications
for regions in nuclear statistical equilibirum.
Furthermore, we reuse the main
CTU hydrodynamics construction and a largely similar ODE integration
scheme, making this method easier to add to existing simulation codes.
It also extends to adaptive mesh refinement with subcycling in a
straightforward manner, avoiding the complications described in
\citep{mccorquodalecolella} needed to fill ghost cells when using
method-of-lines integration.  However, it is restricted to
second-order accuracy overall.  We term this algorithm the
``simplified SDC method''.  We use the \castro\ hydrodynamics code
\citep{castro} for all of our numerical experiments, and all of the
code to reproduce the results in this paper are in the \castro\ github
repository\footnote{\url{https://github.com/amrex-astro/Castro/}}.

@@

We solve the Euler equations for compressible, reacting flow.  For ease
of exposition we describe the one-dimensional case;
multidimensional extensions are a straightforward modification to
include the CTU hydrodynamics scheme.
The system appears as:
\begin{equation}
\frac{\partial \Uc_t}{\partial t}  + \frac{\partial \Fb^\xv (\Uc)}{\partial x} = \Sc(\Uc)
\end{equation}
where
\begin{equation}
\Uc = \left ( \begin{array}{c}
           \rho \\
           \rho X_k \\
           \rho \alpha_l \\
           \rho u \\
           \rho E \\
           \rho e \end{array}\right )
\end{equation}
are the conserved fluid quantities: mass density, $\rho$, velocity,
$u$, specific total energy, $E$, and nuclear species mass fractions,
$X_k$, and auxiliary composition variables, $\alpha_l$.  We also carry
the specific internal energy, $e$, as part of a dual energy
formulation (see \citealt{bryan:1995,wdmergerI}). The mass fractions
are constrained to sum to 1, $\sum_k X_k = 1$, but no such constraint
exists on the auxiliary variables.  The specific total energy relates
to the specific internal energy as $E = e + u^2/2$.  The corresponding
fluxes are
\begin{equation}
\Fb^\xv (\Uc) = \left ( \begin{array}{c}
   \rho u \\
   \rho X_k u \\
   \rho \alpha_l u \\
   \rho u u + p \\
   \rho u E + u p \\
   \rho u e
   \end{array} \right )
\end{equation}
Here the pressure, $p$, enters, and is found via the equation of state.  For a system
where the composition is completely specified by the mass fractions, $X_k$, the equation
of state would take the form:
\begin{equation}
p = p(\rho, X_k, e)
\end{equation}
However, as we'll see shortly, when we use the NSE table, we are using
the results of a much larger network then can be represented by the
mass fractions, and in this case, we rely on the auxiliary composition
variables, $\alpha_l$, to describe the state of the composition, and
our EOS has the form:
\begin{equation}
p = p(\rho, \alpha_l, e)
\end{equation}

We decompose the source terms, $\Sc(\Uc)$, into
hydrodynamical sources, $\Hb$ (like gravity), and reactive sources,
$\Rb$,
\begin{equation}
  \Sc(\Uc) = \Hb(\Uc) + \Rb(\Uc).
\end{equation}
The hydrodynamic sources include the ``pdV'' work for internal energy
as well as gravity
\begin{equation}
  \Hb(\Ub) = \left ( \begin{array}{c}
    0 \\
    0 \\
    0 \\
    \rho g \\
    \rho u g \\
    -p \partial u/\partial x \end{array} \right )
\end{equation}
 The reactive sources take the form:
\begin{equation}
  \label{eq:react_source}
  \Rb(\Uc) = \left ( \begin{array}{c}
     0 \\
     \rho \omegadot(X_k) \\
     \rho \omegadot(\alpha_l) \\
     0 \\
     \rho \dot{S} \\
     \rho \dot{S} 
  \end{array} \right )
\end{equation}
where $\omegadot(X_k)$ is the creation rate for species $k$, $\omegadot(\alpha_l)$ is the creation rate for
auxiliary composition variable $l$, and $\dot{S}$
is the energy generation rate per unit mass.

Sometimes it is preferable to work with the primitive variables,
\begin{equation}
\qb = \left ( \begin{array}{c}
  \rho \\
  X_k \\
  \alpha_l \\
  u \\
  p \\
  (\rho e) \\
\end{array} \right )
\end{equation}
Here, the system appears
as:
\begin{equation}
\qb_t + \Ab^\xv(\qb) \qb_x  = \Sq
\end{equation}
with the matrix $\Ab^\xv$ giving the coefficients of the spatial derivatives
of the primitive variables:
\begin{equation}
\Ab^\xv(\qb) = \left ( \begin{array}{cccccc}
    u & 0 & 0 & \rho & 0 & 0 \\
    0 & u & 0 & 0    & 0 & 0 \\
    0 & 0 & u & 0    & 0 & 0 \\
    0 & 0 & 0 & u    & 1/\rho & 0 \\
    0 & 0 & 0 & \Gamma_1 p & u & 0 \\
    0 & 0 & 0 & \rho h & 0 & u
  \end{array} \right )
\end{equation}
where $\Gamma_1$ is an adiabatic index,
$\Gamma_1 = d\log p/d\log\rho|_s$ at constant entropy.
Note, the primitive state has two thermodynamic quantities, $p$
and $(\rho e)$, to more efficiently handle the general equation of
state in the Riemann solver, as described in \citet{castro}, but
alternate formulations are possible \citep{colellaglaz:1985}.
The source term vector, $\Sq$, can again be decomposed into hydrodynamic
sources (now in terms of the primitive variables) and reaction terms,
\MarginPar{Can we use $\Sc(\qb)$ instead of $\Sq$ and $\Hb(\qb)$ instead of $\Sqhydro$ to keep notation consistent with the conserved variables, since you are already doing that with $\Rb(\qb)$?}
\begin{equation}
  \Sq = \Sqhydro + \Rb(\qb)
\end{equation}
with
\begin{equation}
\label{eq:prim_sources}
\Rb(\qb) = \left ( \begin{array}{c}
     0 \\
     \omegadot(X_k) \\
     \omegadot(\alpha_l) \\
     0 \\
     \Gamma_1 p \sigma \Sdot \\
     \rho \Sdot
   \end{array} \right )
\end{equation}
where
\begin{equation}
\sigma \equiv \frac{\partial p/\partial T |_\rho}{\rho c_p \partial p/\partial \rho |_T}
\end{equation}
and $c_p$ is the specific heat at constant pressure, $c_p = \partial
h/\partial T |_p$.  A derivation of this source for the pressure
equation can be found in \cite{ABNZ:III}.  We note that this source is
algebraically identical to that shown in Eq.~25 of \cite{castro}.

We use the \castro\ hydrodynamics code~\cite{castro}, together with
corner transport upwind piecewise parabolic method
hydrodynamics~\citep{millercolella:2002}.  This is a finite-volume
method that uses characteristic tracing to predict a time-centered
flux through the interfaces of the grid zones.  For pure
hydrodynamics, these time-centered fluxes result in second-order
accurate time integration.  With reactions, we want to couple
hydrodynamics and the reaction sources to second order.  Nuclear
reaction sources are stiff, and need to be integrated using implicit
methods for stabilty.  Operator splitting (e.g., Strang) is traditionally employed
here, and is used as a benchmark for comparison in this paper.

@@

The basic idea of spectral deferred corrections is to express the
update as an integral and divide the time-update into a number of
discrete time nodes.  The integral is then approximated using a
quadrature rule over these time nodes and low order approximations are
used to update the state from one time node to the next.  The method
uses iteration to successively improve the solution and the ultimate
accuracy is determined by the quadrature method used to evaluate the
integral, which is done using an iteratively-lagged solution.

We start by writing our update as:
\begin{equation}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left [ \Advs{\Uc} + \Rb(\Uc) \right ] dt,
\end{equation}
where $\Advs{\Uc} =  -{\partial \Fb^\xv (\Uc)}/{\partial x} + \Hb(\Uc)$.
For the simplified SDC method we explore here, we will make the
advective term piecewise constant in time, using the value at the
midpoint in time, to achieve second-order accuracy, giving us:
\begin{equation}
\label{eq:integral:simplesdc}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left \{ \Adv{\Uc}^{n+1/2} + \Rb(\Uc) \right \} dt
\end{equation}
This is the approach first shown in \citep{SDC-old}.  A
version of this was also implemented in the \maestro\ low-Mach number
hydrodynamics code some time ago.  The compressible version for use in
\castro\ is slightly different due to the need to do some operations
on the conserved variable state and some on the primitive variable
state.  We describe the application of this to the equations of
compressible hydrodynamics below.

@@

  \begin{itemize}
  \item We need an approximation of how much the reactions alone
    changed the primitive variable state over the timestep, which we
    will denote $\Ic_q$.  Since we do not have any information about
    the current timestep in the first iteration,
    we use the value from the last iteration of the previous timestep:
    \begin{equation}
      \Ic^{n+1/2,(0)}_{\qb} = \Ic^{n-1/2,(\smax)}_{\qb}
    \end{equation}
    \MarginPar{changed to 1-based $k$ iteration for aesthetics?  I can switch it back if you prefer ``-1'' to be the starting point instead of ``0''.}
  %% \item Solve the Poisson problem for the initial gravitational potential:
  %%   \begin{equation}
  %%     \nabla^2 \Phi^n = 4\pi G \rho^n
  %%   \end{equation}

@@

    \item predict $\qb$ to the interfaces at $t^{n+1/2}$ using the CTU PPM
      method.  The source terms, $\Sq$, used in the prediction are:
      \begin{equation}
        \Sq = \Sqhydro^{n+1/2} + \Ic_\qb^{n+1/2,(k-1)}
      \end{equation}
      Here we use the iteratively lagged integrals of the primitive variable
      terms accounting only for reactions, $\Ic_\qb^{n+1/2,(k-1)}$, as the
      reactive source.  This is in contrast to Strang-splitting, where no
      explicit reactive source terms are seen by the hydrodynamics update.
      Any hydrodynamic source terms are time-centered
      using the previous iteration:
\MarginPar{Formally both $t^n$ or $t^{n+1/2}$ sources give you 2nd-order since they are CTU predictor source terms.  I've never found this type of iteration for this term to amount to any benefit}
      \begin{equation}
        \Sqhydro^{n+1/2} = \frac{1}{2} \left ( \Sqhydro^n + \Sqhydro^{n+1,(k-1)} \right )
      \end{equation}

@@

>>>>

\newcommand{\nablab}{{\mathbf{\nabla}}}
\newcommand{\Ub}{\mathbf{U}}
\newcommand{\gb}{\mathbf{g}}
\newcommand{\omegadot}{\dot{\omega}}
\newcommand{\Sdot}{\dot{S}}
\newcommand{\ddx}[1]{{\frac{{\partial#1}}{\partial x}}}
\newcommand{\ddxs}[1]{{\frac{{\partial}}{\partial x}}#1}
\newcommand{\ddt}[1]{{\frac{{\partial#1}}{\partial t}}}
\newcommand{\odt}[1]{{\frac{{d#1}}{dt}}}
\newcommand{\divg}[1]{{\nablab \cdot \left (#1\right)}}
\newcommand{\dedr}{\left . {\partial{}e}/{\partial\rho}\right |_{T, X_k}}
\newcommand{\dedrd}{\left . \frac{\partial{}e}{\partial\rho}\right |_{T, X_k}}
\newcommand{\dedX}{\left . {\partial{}e}/{\partial{}X_k} \right |_{\rho, T}}
\newcommand{\dedXd}{\left . \frac{\partial{}e}{\partial{}X_k} \right |_{\rho, T, X_{j,j\ne k}}}
\newcommand{\dedT}{\left . {\partial{}e}/{\partial{}T} \right |_{\rho,X_k}}
\newcommand{\dedTd}{\left . \frac{\partial{}e}{\partial{}T} \right |_{\rho,X_k}}

@@

\author[0000-0001-8401-030X]{M.~Zingale}
\affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
  Stony Brook, NY 11794-3800}

\author[0000-0003-0439-4556]{M.~P.~Katz}
\affiliation{Nvidia Corp}

\author[0000-0003-1791-0265]{A.~J.~Nonaka}
\affiliation{Center for Computational Sciences and Engineering, Lawrence Berkeley National Laboratory, Berkeley, CA  94720}

\author[0000-0002-1530-781X]{Alice Harpole}
\affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
             Stony Brook, NY 11794-3800}

\correspondingauthor{Michael Zingale}
\email{michael.zingale@stonybrook.edu}

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
Reacting astrophysical flows can be challenging to model because of
the difficulty in accurately coupling hydrodynamics and reactions.
This can be particularly acute during explosive burning or at high
temperatures where nuclear statistical equilibrium is established.  We
develop a simplified approach based on the ideas of spectral deferred
corrections (SDC) coupling of explicit hydrodynamics and stiff
reaction sources as an alternative to operator splitting or the more
comprehensive SDC approach we demonstrated previously.  We apply the
new method to some example problems and show how to modify it to work
with a hybrid network consisting of a reaction ODE system and a table
for nuclear statistical equilibrium.  This is all done in the
framework of the \castro\ hydrodynamics code, and all algorithm
implementations are freely available.
\end{abstract}

@@

Modeling astrophysical reacting flows can be challenging because of
the disparity better the nuclear and hydrodynamics timescales.
Reaction networks tend to be stiff, requiring implicit integration
techniques to stably integrate the system.  In contrast, compressible
hydrodynamics flows are limited by the (often much longer)
sound-crossing time over a computational cell and can be solved using
explicit time integration. Traditional methods of coupling
hydrodynamics and reactions used in astrophysics use operator
splitting---each physical process acts on the output of the previous
process in alternating fashion.  This makes it easy to use different
time-integration methods for the different physics, and to build a
simulation code in a modular way.  However, competition between the
different physical processes can cause the coupling to breakdown.
These splitting errors can lead to loss of accuracy and further time
step limitations.

@@

The \castro\ hydrodynamics code \citep{castro,castro_joss} is used for
all of our numerical experiments.  \castro\ is a compressible
(magneto-, radiation-) hydrodynamics code built on the AMReX adaptive
mesh refinement (AMR) framework \citep{amrex_joss}.  \castro\ has been
designed to be performance portable and runs on massively parallel CPU
and GPU architectures \citep{castro_gpu}.  For hydrodynamics, the
corner transport upwind (CTU) \citep{ppmunsplit} method with the
piecewise parabolic method (PPM) \citep{ppm,millercolella:2002} is
used.  \castro\ includes self-gravity, rotation, arbitrary equations
of state and reaction networks, and has been used for modeling X-ray
bursts and different models of thermonuclear, core-collapse, and
pair-instability supernovae.  Recently in \citet{castro_sdc}, we
developed second- and fourth-order accurate methods in space and time
for coupling hydrodynamics and nuclear reaction networks based on
spectral deferred corrections (SDC) methods, and demonstrated the
method in a variety of test problems.

The time-integration approach presented here is considerably simpler than
the SDC method of \citet{castro_sdc}, but allows us to reuse the main CTU
hydrodynamics construction and a largely similar ODE integration
scheme, making this method easier to add to existing simulation codes.
Furthermore, it also extends to adaptive mesh refinement with
subcycling in a straightforward manner, avoiding the complications
described in \citep{mccorquodalecolella} needed to fill ghost cells
when using method-of-lines integration.  However, it is restricted to
second-order accuracy overall.  We term this algorithm the
``simplified SDC method''.  In this paper we describe the overall method, show its extension to NSE,
and demonstrate it on several test problems using \castro.  All of the
code to reproduce the results in this paper are in the \castro\ github
repository\footnote{\url{https://github.com/amrex-astro/Castro/}}.

@@

We solve the Euler equations for compressible, reacting flow.  For ease
of exposition we describe the one-dimensional case;
multidimensional extensions are a straightforward modification to
include the CTU hydrodynamics scheme.
The system appears as:
\begin{align}
\label{eq:conserved_system}
\ddt{\rho} + \ddxs{(\rho u)} &= 0 \\
\ddt{(\rho X_k)} + \ddxs{(\rho X_k u)} &= \rho \omegadot(X_k) \\
\ddt{(\rho \alpha_k)} + \ddxs{(\rho \alpha_k u)} &= \rho \omegadot(\alpha_k) \\
\ddt{(\rho u)} + \ddxs{(\rho u^2 + p)} &= \rho g \\
\ddt{(\rho E)} + \ddxs{(\rho E u + p u)} &= \rho ug + \rho \Sdot 
\end{align}
where $\rho$ is the mass density, $u$ is velocity, $E$ is specific
total energy, $p$ is the pressure, and we carry nuclear species mass
fractions, $X_k$, and auxiliary composition variables, $\alpha_l$.  The specific total
energy relates to the specific internal energy, $e$, as $E = e + u^2/2$,
and we also separately evolve $e$, as part of a dual
energy formulation (see \citealt{bryan:1995,wdmergerI}):
\begin{equation}
\ddt{(\rho e)} + \ddxs{(\rho e u)} + p \ddx{u} = \rho \Sdot
\end{equation}
 The mass fractions are constrained to sum to 1, $\sum_k X_k = 1$, but
 no such constraint exists on the auxiliary variables.  Gravity,
 represented by the gravitational acceleration $g$ appears as a source
 term.  Finally, $\omegadot(X_k)$ is the creation rate for species
 $k$, $\omegadot(\alpha_l)$ is the creation rate for auxiliary
 composition variable $l$, and $\dot{S}$ is the energy generation rate
 per unit mass.  Aside from the $p\partial u/\partial x$ (``pdV'')
 term in the internal energy equation, this system is in conservative
 form with source terms.

To close the system, we need the equation of state.  For a system
where the composition is completely specified by the mass fractions,
$X_k$, the equation of state would take the form:
\begin{equation}
p = p(\rho, X_k, e)
\end{equation}
However, as we'll see shortly, when we use the NSE table, we are using
the results of a much larger network then can be represented by the
mass fractions, and in this case, we rely on the auxiliary composition
variables, $\alpha_l$, to describe the state of the composition, and
our EOS has the form:
\begin{equation}
p = p(\rho, \alpha_l, e)
\end{equation}

For convenience, we'll use $\Uc$ to represent the conserved state
and split the sources up into pure hydrodynamic and reactive parts:
\begin{equation}
  \Uc = \left ( \begin{array}{c}
           \rho \\
           \rho X_k \\
           \rho \alpha_l \\
           \rho u \\
           \rho E \\
           \rho e \end{array}\right )
  \qquad
  \Hb(\Uc) = \left ( \begin{array}{c}
    0 \\
    0 \\
    0 \\
    \rho g \\
    \rho u g \\
    0 \end{array} \right )
  \qquad
  \Rb(\Uc) = \left ( \begin{array}{c}
     0 \\
     \rho \omegadot(X_k) \\
     \rho \omegadot(\alpha_l) \\
     0 \\
     \rho \dot{S} \\
     \rho \dot{S} 
  \end{array} \right )
\end{equation}
We will represent the total source as $\Sc(\Uc)$:
\begin{equation}
  \Sc(\Uc) = \Hb(\Uc) + \Rb(\Uc).
\end{equation}
We note that the internal energy ``pdV'' work is not treated as a source
term but is instead constructed with the hydrodynamical fluxes are computed in the CTU method.

Sometimes it is preferable to work with the primitive variables,
\begin{equation}
\qb = \left ( \begin{array}{c}
  \rho \\
  X_k \\
  \alpha_l \\
  u \\
  p \\
  (\rho e) \\
\end{array} \right )
\end{equation}
Here, the system appears
as:
\begin{equation}
\qb_t + \Ab^\xv(\qb) \qb_x  = \Sc(\qb)
\end{equation}
with the matrix $\Ab^\xv$ giving the coefficients of the spatial derivatives
of the primitive variables:
\begin{equation}
\Ab^\xv(\qb) = \left ( \begin{array}{cccccc}
    u & 0 & 0 & \rho & 0 & 0 \\
    0 & u & 0 & 0    & 0 & 0 \\
    0 & 0 & u & 0    & 0 & 0 \\
    0 & 0 & 0 & u    & 1/\rho & 0 \\
    0 & 0 & 0 & \Gamma_1 p & u & 0 \\
    0 & 0 & 0 & \rho h & 0 & u
  \end{array} \right )
\end{equation}
where $h$ is the specific enthalpy and $\Gamma_1$ is an adiabatic index,
$\Gamma_1 = d\log p/d\log\rho|_s$ at constant entropy.
The CTU+PPM algorithm uses the characteristic wave structure of $\Ab$
to collect the information that makes it to an interface over a timestep
in order to compute the fluxes through the interface.
Note, the primitive state has two thermodynamic quantities, $p$
and $(\rho e)$, to more efficiently handle the general equation of
state in the Riemann solver, as described in \citet{castro}, but
alternate formulations are possible \citep{colellaglaz:1985}.
The source term vector, $\Sc(\qb)$, can again be decomposed into hydrodynamic
sources (now in terms of the primitive variables) and reaction terms,
\begin{equation}
  \Sc(\qb) = \Hb(\qb) + \Rb(\qb)
\end{equation}
with
\begin{equation}
\label{eq:prim_sources}
\Hb(\qb) = \left ( \begin{array}{c}
     0 \\
     0 \\
     0 \\
     g \\
     0 \\
     0 \\
   \end{array} \right )
\qquad
\Rb(\qb) = \left ( \begin{array}{c}
     0 \\
     \omegadot(X_k) \\
     \omegadot(\alpha_l) \\
     0 \\
     \Gamma_1 p \sigma \Sdot \\
     \rho \Sdot
   \end{array} \right )
\end{equation}
where
\begin{equation}
\sigma \equiv \frac{\partial p/\partial T |_\rho}{\rho c_p \partial p/\partial \rho |_T}
\end{equation}
and $c_p$ is the specific heat at constant pressure, $c_p = \partial
h/\partial T |_p$.  A derivation of this source for the pressure
equation can be found in \cite{ABNZ:III}.  We note that this source is
algebraically identical to that shown in Eq.~25 of \cite{castro}.

The CTU+PPM method for hydrodynamics is second-order accurate in space
and time.  We want to couple the reactive sources to the hydrodynamics
to be second-order in time as well.  As discussed above, nuclear
reaction sources are stiff, and need to be integrated using implicit
methods for stabilty.  Operator splitting (e.g., Strang) is
traditionally employed here, and is used as a benchmark for comparison
in this paper.  We'll discuss this traditional approach next before moving
on to our new time-coupling method.

@@

The basic idea of spectral deferred corrections \MarginPar{ref} is to express the
update as an integral and divide the time-update into a number of
discrete time nodes.  The integral is then approximated using a
quadrature rule over these time nodes and low order approximations are
used to update the state from one time node to the next.  The method
uses iteration to successively improve the solution and the ultimate
accuracy is determined by the quadrature method used to evaluate the
integral, which is done using an iteratively-lagged solution.  This is the
classic approach that we implemented in \cite{castro_sdc}.

Our simplified-SDC approach is based on the method described in \citet{SDC-old} and
a similar implementation in the \maestro\ simulation code.
We again start by considering the update
in integral form:
\begin{equation}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left [ \Advs{\Uc} + \Rb(\Uc) \right ] dt,
\end{equation}
where $\Advs{\Uc}$ includes all terms in system
\ref{eq:conserved_system} except for the reactive source terms,
$\Rb(\Ub)$ (i.e., it includes the flux divergences, ``pdV'' term for
internal energy, and the pure hydrodynamical sources).  In this approach, we 
approximate the advective term
piecewise constant in time, using the value at the midpoint in time,
to achieve second-order accuracy, giving us:
\begin{equation}
\label{eq:integral:simplesdc}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left \{ \Adv{\Uc}^{n+1/2} + \Rb(\Uc) \right \} dt
\end{equation}
The \castro\ version of the simplified-SDC algorithm
differs from the previous versions due to the need to do some operations
on the conserved variable state and some on the primitive variable
state.

The basic time update algorithm proceeds as:

@@

  \begin{itemize}
  \item We need an approximation of how much the reactions alone
    changed the primitive variable state over the timestep, which we
    will denote $\Ic_q$ (this is basically an approximation of
    $\Rb(q)$).

    Since we do not have any information about the current
    timestep in the first iteration, we use the value from the last
    iteration of the previous timestep:
    \begin{equation}
      \Ic^{n+1/2,(0)}_{\qb} = \Ic^{n-1/2,(\smax)}_{\qb}
    \end{equation}
  %% \item Solve the Poisson problem for the initial gravitational potential:
  %%   \begin{equation}
  %%     \nabla^2 \Phi^n = 4\pi G \rho^n
  %%   \end{equation}

@@

    \item predict $\qb$ to the interfaces at $t^{n+1/2}$ using the CTU PPM
      method.  The source terms, $\Sq$, used in the prediction are:
      \begin{equation}
        \Sc(\qb) = \Hb(\qb) + \Ic_\qb^{n+1/2,(k-1)}
      \end{equation}
      Here we use the iteratively lagged integrals of the primitive variable
      terms accounting only for reactions, $\Ic_\qb^{n+1/2,(k-1)}$, as the
      reactive source.  This is in contrast to Strang-splitting, where no
      explicit reactive source terms are seen by the hydrodynamics update.
      Any hydrodynamic source terms are time-centered
      using the previous iteration:
\MarginPar{Formally both $t^n$ or $t^{n+1/2}$ sources give you 2nd-order since they are CTU predictor source terms.  I've never found this type of iteration for this term to amount to any benefit}
      \begin{equation}
        \Sqhydro^{n+1/2} = \frac{1}{2} \left ( \Sqhydro^n + \Sqhydro^{n+1,(k-1)} \right )
      \end{equation}

@@

Do this in 1-d
Show the amount of work needed for Strang vs. SDC
